---
title: "Background extension"
output:
  html_document:
    toc: yes
    fig_width: 10
    fig_height: 5
---

## Load libraries

Mizer 2.0.1 updated at 23/05/2020 or later version is needed to use the background extension. 

A. First and best option: download the most updated version of the package directly from GitHub using devtools::install_github("sizespectrum/mizer") 

B. Second option: merge upstream master with your branch and upload code from local repository 

```{r message=FALSE, warning=FALSE}

rm(list=ls())

# option A
# devtools::install_github("sizespectrum/mizer") # run only once
library(mizer)

# option 2 
# library(Jmisc)
# path<-"/Users/nov017/R-projects/Mizer-trait/R"
# sourceAll(path)

```

## Set up multispecies model

All parameters are made up

```{r message=FALSE, warning=FALSE}

species_params = data.frame(species = c("pelagic", "benthic"), w_inf = c(100, 200), k_vb =c(0.2, 0.2), w_min = c(0.001, 0.001))

# available PP to species 
species_params$interaction_resource<-c(1,0.2)

# create the param object
# note: some def values are different from old mizer (e.g. kappa)
params<-newMultispeciesParams(species_params, no_w = 200, kappa = 0.001)
params@resource_params

# run projections:
sim<-project(params, t_max = 100, effort = 0)
plot(sim)

```

## Add a second background spectrum to the Param object

To create a new component, the background resource spectrum, you need to use setComponent(). See **extension.r** in Mizer package for a detailed description of this function and its arguments 

In brief, setComponent (params, component, initial_value, dynamics_fun, encounter_fun, mort_fun, component_params)

params = A MizerParams object  
component = Name of the component  
initial_value = Initial value of the component  
dynamics_fun = Name of function to calculate value at the next time step  
encounter_fun = Name of function to calculate contribution to encounter rate. Optional  
mort_fun = Name of function to calculate contribution to the mortality rate. Optional  
component_params = Named list of parameters needed by the component functions. Optional    

### How it works 

We need to specify our own functions (dynamics, encounter and mortality) for each background resource component we'd like to add. 

This can be done in 6 steps:   

1. Specify the background dynamics function  
dynamics_fun = background_bb_semichemostat, which is the same as resource_semichemostat() in **resource_dynamics.r**, but uses newly defined rr_pp (here called rate) and cc_pp (here called capacity), and it includes the mortality function  

```{r}

background_bb_semichemostat <- function(params, n_other, rates, dt, ...) {
    
    # locate where (which params slot) the parameters needed for this function are stored 
    component <- "background_bb"
    c <- params@other_params[[component]] 
    
    # name of interaction parameter for this component in species_params
    interaction_component <- paste0("interaction_", component)
    interaction <- params@species_params[[interaction_component]]
    
    # Step 3: specify the background mortality function
    mort <- as.vector(interaction  %*% rates$pred_rate)
    
    # specify the background dynamics function. This is the same as resource_semichemostat() - see above
    tmp <- c$rate * c$capacity / (c$rate + mort) 
    return(tmp - (tmp - n_other[[component]]) * exp(-(c$rate + mort) * dt))
}

```

2. Specify the encounter function  
encounter_fun = background_bb_encounter, which is same as mizerEncounter() in **project_methods.r**, but now n_pp is the sum of the 2 background spectra  

```{r}

background_bb_encounter <- function(params, n, n_pp, n_other, ...) {
    mizerEncounter(params, n = n, n_pp = n_pp + n_other$background_bb, ...)
}

```

3. Specify the mortality function  
mort_fun = mort, which is the same as mizerResourceMort() in **project_methods.r**, but uses different interaction values (interaction_resource) and pred_rate. Mortality in this case is specified as part of background_bb_semichemostat(). For any other extensions you'd need something like: 

```{r}

# background_bb_mortality<-...mizerMortality()...

```

4. Set up parameters   
Calcualate rate and capacity used in background_bb_semichemostat() and given kappa, lambda etc. for the bb spectrum

```{r}

kappa <- 0.001 
lambda <- 2.05 
r_pp <- 10
n <- 0.666 
w_pp_cutoff <- 10
rate <- r_pp * params@w_full^(n - 1)
capacity <- kappa * params@w_full^(-lambda)
capacity[params@w_full > w_pp_cutoff] <- 0
w_pp_min <- 1e-6 # set minimum size of new resource 
capacity[params@w_full > w_pp_cutoff] <- 0

# Set interaction species-background_bb - i.e. proportion available to each species.  
# this new species param column needs to be named as "interaction_" + component name
params@species_params$interaction_background_bb <- c(0,0.8)

```

5. Add components   
Add the background bb resource spectrum to the params object using setComponent()

```{r}

params2 <- setComponent(params = params, component = "background_bb",
                        initial_value = params@initial_n_pp, 
                        dynamics_fun =  "background_bb_semichemostat",
                        component_params = list(rate = rate, 
                                                capacity = capacity))

```

6. Update the encounter function

```{r}

params2 <- setRateFunction(params2, "Encounter", "background_bb_encounter")

```

7. Update the mortality function   
Mortality in this case is specified as part of background_bb_semichemostat() so you don't need to update this function. For any other extensions you'd need to run the line below: 

```{r}

# params2<-setRateFunction(params2, "Mortality", "background_bb_mortality")

```

## Run project() using the newly defined Param object 

Params now includes a pp and a bb background spectrum 

```{r message=FALSE}

sim2 <- project(params2, t_max = 100, effort = 0) 
plot(sim2)

# check the community 
# sim2@n_other[10,] # n background_bb at time step 10 

```

## Plot community with additional resource

"to plot the results you will need to write your own plot function. You can of course use the built-in plot functions to get a ggplot2 object to which you then only need to add the extra stuff for your new components".
