---
title: "therMizer extension"
output:
  html_document:
    toc: yes
    fig_width: 10
    fig_height: 5
---

## Load libraries

Mizer 2.0.1 updated at 23/05/2020 or later version is needed to use the therMizer extension. 

A. First and best option: download the most updated version of the package directly from GitHub using devtools::install_github("sizespectrum/mizer") 

B. Second option: merge upstream master with your branch and upload code from local repository 

```{r message=FALSE, warning=FALSE}

rm(list=ls())

# option A
# devtools::install_github("sizespectrum/mizer") # run only once
library(mizer)

# option 2 
# library(Jmisc)
# path<-"/Users/nov017/R-projects/Mizer-trait/R"
# sourceAll(path)

```

## Set up multispecies model

All parameters are made up

```{r message=FALSE, warning=FALSE}

species_params = data.frame(species = c("pelagic", "benthic"), w_inf = c(100, 200), k_vb =c(0.2, 0.2), w_min = c(0.001, 0.001), temp_min = c(15, 10), temp_max = c(25, 20))

# available PP to species 
species_params$interaction_resource<-c(1,0.2)

# create the param object
# note: some def values are different from old mizer (e.g. kappa)
params<-newMultispeciesParams(species_params, no_w = 200, kappa = 0.001)
params@resource_params

# run projections:
sim<-project(params, t_max = 100, effort = 0)
plot(sim)

```

## Add therMizer functions to the Param object

To create a new component, the background resource spectrum, you need to use setComponent(). See **extension.r** in Mizer package for a detailed description of this function and its arguments 

In brief, setComponent (params, component, initial_value, dynamics_fun, encounter_fun, mort_fun, component_params)

params = A MizerParams object  
component = Name of the component  
initial_value = Initial value of the component  
dynamics_fun = Name of function to calculate value at the next time step  
encounter_fun = Name of function to calculate contribution to encounter rate. Optional  
mort_fun = Name of function to calculate contribution to the mortality rate. Optional  
component_params = Named list of parameters needed by the component functions. Optional    

### How it works 

We need to specify our own functions (dynamics, encounter, and mortality). 

This can be done in 6 steps:   
1. Calculate the additional paramaters needed 
Value to scale temperature effect on encounter rate
```{r}
for (indv in seq(1:length(params@species_params$temp_min))){
			temperature <- seq(params@species_params$temp_min[indv], params@species_params$temp_max[indv], by=0.1)
			all_possible_encount_scale_values <- (temperature) * (temperature - params@species_params$temp_min[indv]) * (params@species_params$temp_max[indv] - temperature)
			params@species_params$encount_scale[indv] <- max(all_possible_encount_scale_values)
			}
```

Values to scale temperature effect on metabolism
```{r}
    params@species_params$metab_min <- (exp(25.22 - (0.63/((8.62e-5)*(273+params@species_params$temp_min))))) 

    min_metab_value <- (exp(25.22 - (0.63/((8.62e-5)*(273+params@species_params$temp_min)))))
		max_metab_value <- (exp(25.22 - (0.63/((8.62e-5)*(273+params@species_params$temp_max)))))
		params@species_params$metab_range <- max_metab_value - min_metab_value
```

2. Specify the encounter function  
encounter_fun = therMizer_encounter, which is same as mizerEncounter() in **project_methods.r**, but now with temperature included.  It is modeled as a scaled simple polynomial rate equation  

```{r}

therMizer_encounter <- function(params, n, n_pp, n_other, ocean_temp, ...) {
      # Calculate scaled temperature effect
      unscaled_temp_effect <- (ocean_temp) * (ocean_temp - params@species_params$temp_min) * (params@species_params$temp_max - ocean_temp)
      
      scaled_temp_effect <- unscaled_temp_effect / params@species_params$encount_scale
      
      mizerEncounter(params, n = n, n_pp = n_pp, n_other = n_other, ...)*scaled_temp_effect
}

```

3. Specify the energy available for growth and reproduction function  
For any other extensions you'd need something like: 

```{r}

therMizer_mizerEReproAndGrowth <- function(params, n, n_pp, n_other, encounter,
                                 feeding_level, ocean_temp, ...) {

      unscaled_temp_effect <- (exp(25.22 - (0.63/((8.62e-5)*(273+ocean_temp)))))
			temp_effect_metabolism <- (unscaled_temp_effect - params@species_params$metab_min) / params@species_params$metab_range
  
    sweep((1 - feeding_level) * encounter, 1,
               params@species_params$alpha, "*", check.margin = FALSE) - 
        params@metab*temp_effect_metabolism
}

```

4. Set up parameters   


```{r}

# kappa <- 0.001 
# lambda <- 2.05 
# r_pp <- 10
# n <- 0.666 
# w_pp_cutoff <- 10
# rate <- r_pp * params@w_full^(n - 1)
# capacity <- kappa * params@w_full^(-lambda)
# capacity[params@w_full > w_pp_cutoff] <- 0
# w_pp_min <- 1e-6 # set minimum size of new resource 
# capacity[params@w_full > w_pp_cutoff] <- 0
# 
# # Set interaction species-background_bb - i.e. proportion available to each species.  
# # this new species param column needs to be named as "interaction_" + component name
# params@species_params$interaction_background_bb <- c(0,0.8)

```

5. Add components   
Add the background bb resource spectrum to the params object using setComponent()

```{r}

# params2 <- setComponent(params = params, component = "background_bb",
                        # initial_value = params@initial_n_pp, 
                        # dynamics_fun =  "background_bb_semichemostat",
                        # component_params = list(rate = rate, 
                        #                         capacity = capacity))

```

6. Update the encounter function

```{r}

params <- setRateFunction(params, "Encounter", "therMizer_encounter")
params <- setRateFunction(params, "EReproAndGrowth", "therMizer_mizerEReproAndGrowth")
```

7. Update the mortality function   
Mortality in this case is specified as part of background_bb_semichemostat() so you don't need to update this function. For any other extensions you'd need to run the line below: 

```{r}

# params2<-setRateFunction(params2, "Mortality", "background_bb_mortality")

```

## Run project() using the newly defined Param object 

Params now includes a pp and a bb background spectrum 

```{r message=FALSE}

sim2 <- project(params, t_max = 100, effort = 0, ocean_temp = 17) 
plot(sim2)

# check the community 
# sim2@n_other[10,] # n background_bb at time step 10 

```

## Plot community with additional resource

"to plot the results you will need to write your own plot function. You can of course use the built-in plot functions to get a ggplot2 object to which you then only need to add the extra stuff for your new components".
